Monitoring Time Series With Missing Values:\\a Deep Probabilistic Approach

### 2022-02-15 paper.tex
in progress

The forecasting is accomplished by applying model $\mathcal{M}_\theta$
parameterized by parameters $\theta$ to the observed time
series:
\begin{equation}
\mathcal{F}_{tk} = \mathcal{M}_\theta(X_{1:t})
\end{equation}
The machine learning task is to devise $\theta^*$
that gives the best forecast, in terms of a certain loss
function. A natural loss in the probabilistic setting is the
average negative log likelihood of $\theta$ given a training
data set $\mathcal{X}$ of multiple time series:
\begin{equation}
	\theta^* = \arg\min_\theta \mathbb{E}_{X \in \mathcal{X},t
	\in 1 ... T-k}\left[ -\log \Pr(X_{t+1:t+k}|M_\theta(X_{1:t}))\right]
\end{equation}
When the model is differentiable by $\theta$, the task is
usually accomplished by performing a stochastic gradient loss
minimization.

@@

\paragraph{Input and Output} This architecture normally accepts
data vectors and outputs vectors of predictions of means and
standard deviations (the output is twice as wide as the input).
Then, to generate a forecast, one samples from
$$\mathcal{N}(\mu, \sigma)$$ for each component, and feeds the
sample as the input to the next time step.

\paragraph{Training} The model is trained to maximize
probability of prediction. In the most basic case, a single step
is predicted for each time step in the series. If a series has
10 steps, then steps 1 .. 9 are the input, and the steps 2 .. 10
are the truth. The network is trained to minimize negative log
probability of true observations given predictions under
variational approximation of product of Gaussians.

@@

\begin{enumerate}
\item The input, as well as the output, is a vector of means followed
by standard deviations. If the data has 5 components, the input
will be 10-dimensional. For observed data --- measurements
present at the current time step --- the standard deviation is
zero. For missing data the input is the mean and the standard deviation
as predicted from the preceding time steps.

@@

The ability of probabilistic forecasting with uncertainty, in
the form of a multivariate Gaussian distributions, far into the
future, opens opportunity for application for more robust
novelty detection approaches. Instead of detecting novelty based
on log probability of observations given predictions from the
past, which is prone to false positives due to observation
noise, novelties can be detected and analysed by comparing
predictions of the same time point from different points in the
past. In this case, KL-divergence between predictions provide a
theoretically sound and robust mechanism for detection of
anomalies, and is in particular relevant for monitoring of large
operation environments with high dimensionality of time series
and occasional missing values and heteroskedastic noise.

>>>>

The forecasting is accomplished by applying model $\mathcal{M}_\theta$
parameterized by parameters $\theta$ to the observed time
series:
\begin{equation}
\mathcal{F}_{tk} = \mathcal{M}_\theta(X_{1:t})
\end{equation}
The machine learning task is to devise $\theta^*$
that gives the best forecast, in terms of a certain loss
function. A natural loss in the probabilistic setting is the
average negative log likelihood of $\theta$ given a training
data set $\mathcal{X}$ of multiple time series:
\begin{equation}
	\theta^* = \arg\min_\theta \mathbb{E}_{X \in \mathcal{X},t
	\in 1 ... T-k}\left[ -\log \Pr(X_{t+1:t+k}|M_\theta(X_{1:t}))\right]
	\label{eqn:theta-star}
\end{equation}
When the model is differentiable by $\theta$, the task is
usually accomplished by performing a stochastic gradient loss
minimization.

@@

\paragraph{Input and Output} This architecture normally accepts
observation vectors and outputs vectors of distribution
parameters for the belief distribution of the observations at
the next time step. In the case of frequently used normal
distributions, the output vector consists of means $\mu$ and
standard deviations $\sigma$ of all dimensions and is twice as
wide as the input.  This allows to also assess uncertainty
multiple steps into the future: one can repeatedly sample from
the observation belief distribution at each future step, and
feed the sample as the input to the following step.

\paragraph{Training} The model is trained to maximize
probability of prediction. In the most basic case, a single step
is predicted for each time step in the series. If a series has
10 steps, then steps $1 ... 9$ are the input, and the steps $2
... 10$ are the truth. Following~\eqref{eqn:theta-star}, the
network is trained to minimize negative log probability of true
observations given the predicted belief distributions.

@@

\begin{enumerate}
\item The input, as well as the output, is a vector of
distribution parameters. For simplicity, we will confine further
discussion to the case of independent normal belief
distributions for each component; however, other distribution
shapes can be used. For the independent normal
distributions, the distribution parameter vector consists of the
means followed by the standard deviations. If the data has 5
components, the input will be 10-dimensional. For observed data
--- measurements present at the current time step --- the
standard deviation is zero. For missing data the input is the
mean and the standard deviation as predicted from the preceding
time steps.

@@

The ability of probabilistic forecasting with uncertainty, in
the form of multivariate normal distributions, far into the
future, opens opportunity for application to more robust
novelty detection approaches. Instead of detecting novelty based
on log probability of observations given predictions from the
past, which is prone to false positives due to observation
noise, novelties can be detected and analysed by comparing
predictions of the same time point from different points in the
past. In this case, KL-divergence between predictions provide a
theoretically sound and robust mechanism for detection of
anomalies, and is in particular relevant for monitoring of large
operation environments with high dimensionality of time series
and occasional missing values and heteroskedastic noise.
